Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-11-01 manuscript/climate_sensitivity_SI.Rmd
worked on references/discussion, minor text changes

---
title: "Supplementary Information: \nTree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest"
author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Erika Gonzalez-Akre, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
output:
  pdf_document:
    fig_caption: yes
biblio-style: apalike
csl: apa.csl
header-includes:
- \usepackage{booktabs}
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
---

@@

```{r Figure S1, echo=FALSE, out.width='100%', fig.cap = "Figure S1: Map of ForestGEO plot", fig.pos='ht'}
  knitr::include_graphics("tables_figures/figureS1_ForestGEO_plot.jpg")
```

\newpage
*how do we want to present Table S4? Would it be better as an image of an excel file, since it's so large? Did we want to keep all coefficients here?*
```{r, Table S4, eval=TRUE, echo=FALSE, message=FALSE}
tableS4 <- read.csv("tables_figures/tableS4_top_models_dAIC.csv", stringsAsFactors = FALSE)
kable(tableS4, booktabs=TRUE, format = "latex", caption = "Table S4: Top model variations for each drought scenario, with dAICc values <= 2") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

![Figure S2: Location of cored trees](tables_figures/figureS2_location_cored_trees.png){width=500px}

\newpage
![Figure S3: Traits with TWI](tables_figures/figureS3_traits_with_twi.png)
(see Issue #32)

>>>>

---
title: "Supplementary Information: \nTree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest"
author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Erika Gonzalez-Akre, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
output:
  pdf_document:
    fig_caption: yes
biblio-style: apalike
csl: apa.csl
header-includes:
- \usepackage{booktabs}
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{caption}
- \captionsetup[table]{font=small}
- \captionsetup[figure]{font=small}
- \captionsetup[table]{labelformat=empty}
- \captionsetup[figure]{labelformat=empty}
---

@@

\newpage

```{r Table S4, eval=TRUE, echo=FALSE, message=FALSE}
tableS5 <- read.csv("tables_figures/tableS5_tested_traits_height.csv", stringsAsFactors = FALSE)

tableS5 <- tableS5[,-1]
names(tableS5) <- c("variable", "model", "coefficient", "p-value")
tableS5[,c(3)] <- round(tableS5[,c(3)], 2)

tableS5$variable <- c("WD", "LMA", "ring porosity", "PLA", "TLP")
tableS5$model <- c("WD~ln[H]","LMA~ln[H]","ring porosity~ln[H]", "PLA~ln[H]", "PLA~ln[H]")

kable(tableS5, booktabs=TRUE, format = "latex", caption = "Table S4. Correlation of species' traits with tree height across all individuals in the ForestGEO plot") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

\newpage
![Figure S1: Map of ForestGEO plot showing TWI and location of cored trees](tables_figures/figureS2_location_cored_trees.png){width=500px}

### 2019-11-01 manuscript/climate_sensitivity_draft.Rmd
worked on references/discussion, minor text changes

Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironemnt, and species traits (Table 2). The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998]. From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below). Census data, which were last updated in 2018 (**the main census data was taken in 2018, but otherwise we had the final update from Suzanne in 2019. Did you want 2019 here?**), are available through the ForestGEO data portal (www.forestgeo.si.edu).

@@

We analyzed tree-ring data from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019]. Cores were obtained in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer. In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals of DBH $\ge$ 10cm [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).  *Ryan submitted the data to ITRDB but I don't think its posted yet. We should also cite GitHub/Zenodo here. I'll come back to that.*

@@

Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R. Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006]. TWI calculation depends on an input of a digital elevation model (DEM), and from this yields a quantitative assessment defined by how "wet" an area is, based on areas where run-off is more likely. From our observations in the plot, the calculation of TWI performed comparatively better at categorizing wet areas than the calculation of a distance matrix from a stream shapefile.

@@

We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance. In part because of the large sample size (**n=#** trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S5). However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the cases of $WD$ $LMA$, which were not useful predictors of $R$. Furthermore, although correlations were statistically signifcant, trait variation within each height class overwhelmed any vertical trends (Fig. 2e-f).

@@

Beyond microenvironment, multiple tree characteristics covary with height. To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including leaf area and crown volume, rooting volume, sapwood area, and transpiration (e.g., [@andersonteixeira_size-related_2015;  **REFS--KAT: Sperry scaling model; roots reference; BAAD database?**)]. These work in combination to shape whole-plant hydraulic function (**REFS-KAT: Sperry scaling model, check McDowell scaling paper, Couvreur et al. 2018; others?**)[@couvreur_water_2018], and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees. However, height is arguably the most biophysically meaningful tree-size variable shaping drought response. Beyond whole-plant characteristics, leaf traits vary with height on the tree--sometimes quite dramatically, with upper-canopy leaves tending to be smaller and thicker (higher LMA), with more negative $P50$ and lower hydraulic conductivity [@couvreur_water_2018; @koike_leaf_2001]*[McDowell et al. 2011](https://www.springer.com/gp/book/9789400712416)*. Our analysis focuses on species-level comparisons and does not characterize the role of variation with height. On an individual level, average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits--assuming trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing drought resistance with height. In not accounting for this, our analysis is most likely biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$. Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses [@couvreur_water_2018].

The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time (**verify- Neil, Alan**). Concerted measurement of leaf hydraulic traits of emerging importance (**REFS--KAT/NOBBY/LAWREN**) allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients (**REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**... (**REFS**). It is scientifically exciting in that it indicates that these traits, which can be measured relatively easily $PLA_{dry}$ and $\pi_{tlp}$ (**REFS--KAT/NOBBY/LAWREN**), hold promise for predicting drought growth responses across species. The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study; **REFS**), this becomes difficult to impossible for diverse tropical forests, where linking hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS--KAT/NOBBY/LAWREN**).

@@

As climate change drives increasing drought in many of the world's forests (*REFS-KAT*), the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here. Large trees have been suffering disrpoportionately in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we here show that this is primarily driven by their height, with some contributions from canopy position. The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable. Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests (Bretfield et al.), and species traits often shift as forests age, with early successional species tending to have lower wood densities and higher hydraulic conductivities that facilitate rapid growth (**REFS in Bretfield-beginning of discussion**). The later successional species at our site (*Fagus grandifolia*, ... ) have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for getting at the very signficant question of whether old-growth forests are more vulnerable to drought, and to which types of drought. In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.

>>>>

Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironemnt, and species traits (Table 2). The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998]. From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below). Census data, which were last updated in 2019, are available through the ForestGEO data portal (www.forestgeo.si.edu).

@@

We analyzed tree-ring data from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1). Cores were obtained in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer. In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals of DBH $\ge$ 10cm [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).  *Ryan submitted the data to ITRDB but I don't think its posted yet. We should also cite GitHub/Zenodo here. I'll come back to that.*

@@

Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R (Figure S1). Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006]. TWI calculation depends on an input of a digital elevation model (DEM), and from this yields a quantitative assessment defined by how "wet" an area is, based on areas where run-off is more likely. From our observations in the plot, the calculation of TWI performed comparatively better at categorizing wet areas than the calculation of a distance matrix from a stream shapefile.

@@

We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance. In part because of the large sample size (**n=#** trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4). However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the cases of $WD$ $LMA$, which were not useful predictors of $R$. Furthermore, although correlations were statistically signifcant, trait variation within each height class overwhelmed any vertical trends (Fig. 2e-f).

@@

Beyond microenvironment, multiple tree characteristics covary with height. To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including leaf area and crown volume, fine root biomass and rooting volume, sapwood area, and transpiration (e.g., [@andersonteixeira_size-related_2015]; *Chen et al. 2018: DOI: 10.1111/ele.13193.*). These work in combination to shape whole-plant hydraulic function (**REFS-KAT: check McDowell scaling paper,others?** *Sperry et al. 2012 (doi: 10.1111/j.1365-2435.2012.02022.x)*)[@couvreur_water_2018], and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees. However, height is arguably the most biophysically meaningful tree-size variable shaping drought response. Beyond whole-plant characteristics, leaf traits vary with height on the tree--sometimes quite dramatically, with upper-canopy leaves tending to be smaller and thicker (higher LMA), with more negative $P50$ and lower hydraulic conductivity [@couvreur_water_2018; @koike_leaf_2001]*[McDowell et al. 2011](https://www.springer.com/gp/book/9789400712416)*. Our analysis focuses on species-level comparisons and does not characterize the role of variation with height. On an individual level, average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits--assuming trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing drought resistance with height. In not accounting for this, our analysis is most likely biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$. Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses [@couvreur_water_2018].

The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time (**verify- Neil, Alan**). Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019](**MORE REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**... (**REFS**). It is scientifically exciting in that it indicates that $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species. The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for diverse tropical forests, where linking hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--KAT/NOBBY/LAWREN**).

@@

As climate change drives increasing drought in many of the world's forests (*REFS-KAT*), the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here. Large trees have been suffering disrpoportionately in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we here show that this is primarily driven by their height, with some contributions from canopy position. The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable. Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests (*Bretfield et al.*), and species traits often shift as forests age, with early successional species tending to have lower wood densities and higher hydraulic conductivities that facilitate rapid growth (**REFS in Bretfield-beginning of discussion**). The later successional species at our site (*Fagus grandifolia*, ... ) have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for getting at the very signficant question of whether old-growth forests are more vulnerable to drought, and to which types of drought. In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.
